A statistical mechanics approach to the simulation of extreme fluctuations of the time-averaged drag acting on an obstacle immersed in a turbulent flow

### 2018-12-29 draft_p2.tex
Section sur GKTL. Maque conclusion de la section

Figure~\ref{fig:instant_256_20} can be better understood by visualizing the set of the last resampled trajectories. 
One can see that, ultimately, trajectories have been resampled from the same initial trajectory.
We explain this phenonemon as follows.
Because the flow dynamics hae a memory, it takes a certain time before a resampled trajectory separate from its parent.
This is clearly visible on figure~\ref{fig:instant_256_20}.
In our case, this memory originates from the deterministic nature of the dynamics, which implies that two trajectories starting from infinitesimally close starting point will overlap over a timescale called
the <<Lyapunov timesacle>>~\cite{lyapunovRef}.
However, we observe that the duration of extreme drag fluctuations is typically 5 times smaller that the Lyapunov timescale of the drag proces.
This means that the resampling of a trajectory based on the value of the drag at $t^{\star}$ cannot lead to larger drag values for $t^{\star} \leq t \leq \tau_L$.
For $t > \tau_L$, the drag process has no memory of the drag fluctuations on which the resampling was based and the probability of observing an extreme fluctuation over $\tau_l \leq t \leq T_a$ is very low.
The difference between the typical duration of drag fluctuations $\tau_c$ and the Lyapunov timescale $\tau_0$ is linked to the convective nature of the flow.
Indeed, fluid structures responsible for a particular value of the drag on the obstacle at a time $t^{\star}$ are swept away by the mean flow over a timescale $\tau_c$.

@@

Figure~\ref{fig:illustration_IS} illustrates the importance sampling performed by the \ac{gktl} algorithm.
illustrates the amplitude of the fluctuations of $F_{T_a}$ corresponding to the sampled trajectories for each three experiments.
More precisely, it illustrates the estimate of the biased drag \ac{pdf} $\rho_k$ in each case:
\begin{multline}
  \label{eq:estimate_biased_measure}
  \rho_k(F) = \mathbb{E}_{\set{P}_k}[\delta(F_{T_a}[\{\mathbf{x}\}_{0\leq t \leq T_a}]-F)] \\
  \approx \frac{1}{N_c}\sum_{j=1}^{N_c}\delta(F^{(j)}_{T_{a}}[\{\mathbf{x}\}_{0\leq t \leq T_a}] - F).
\end{multline}
Moreover, figure~\ref{fig:illustration_IS_drag} displays an estimate of the unbiased drag \ac{pdf} $\rho_0(F) = \mathbb{E}_{\set{P}_0}[\delta(F_{T_a} - F)]$.
Note that this figure is analogous to figure~\ref{fig:illstration_importance_sampling} on page \pageref{fig:illstration_importance_sampling}.
One can see that, as $k$ is increased, the biased \ac{pdf} is shifted towards extreme values.
However, the experiment with $k=0.03$ does not seem to increase the sampling of rare event with respect to the experiment with $k=0.025$.
Actually, the shape of the estimate of the biased \ac{pdf} shows that the reconstructed trajectories concentrate around a single value of the fluctuation $F'_{T_a} = F_{T_a}-\overline{F_{T_a}} \approx 5\sigma_{T_a}$, where $\overline{F_{T_a}}$ and $\sigma_{T_a}$ denote the average and standard deviation of $F_{T_a}$, respectively.
This suggests a strong overlap of trajectories in the reconstructed ensemble, which we address next.

\paragraph{}
Figure~\ref{fig:nbIndep} illustrates the overlap of trajectories in the reconstructed ensemble.
Along the X-axis, we represent the $M = T_a / \tau$ iterations of the evolution/selection procedure.
For each iteration $m$ we plot the number of independent trajectories, \textit{i.e.} the number of trajectories that exhibit different paths $\{X(t)\}_{(m-1)\tau \leq t \leq m\tau}$ over the interval $t\in [(m-1)\tau;m\tau]$.
In practice we trace back the history of the trajectories backward in time starting from the final time $T_a = M\tau$, in a way very similar to the \emph{Method 1} presented in section~\ref{sec:reconstruction_practice}.
Figure~\ref{fig:nbIndep} illustrates that, at early times, the vast majority of the trajectories in the reconstructed ensemble overlap.
This is also illustrated, in a different way, by the red trajectories in figure~\ref{fig:schema_algo_b}.
Say $N$ trajectories of the reconstructed ensemble overlap over the interval $[0;m\tau]$, with $1\leq m \leq M$.
Among these $N$ trajectories, only one trajectory $j$ actually originates from $t=0$.
Indeed, the $N-1$ others have been discarded following the selection stage of a given iteration $\tilde{m}\geq m$, and replaced by a clone of trajectory $j$ over the interval $[0;\tilde{m}\tau]$.
In the following we call \emph{fully independent trajectories} the trajectories that do not overlap at all over $[0;T_a]$, \textit{i.e.} trajectories that do not overlap over the first period $[0;\tau]$.
From figure~\ref{fig:nbIndep} one can see that the number of fully independent trajectories is very small compared to the total number of trajectories $N_c$.

\paragraph{}
In this context, can the \ac{gktl} yield more \emph{independent} events than a direct sampling based on a long simulation ?
If yes, how much more ?
\begin{table}
  \centering
  \subfloat[$k=0.02$]{
  \begin{tabular}{l|ccccc}
    Overlap & $\sigma$ & $2\sigma$ & $3\sigma$ & $4\sigma$ & $5\sigma$ \\
    \hline
    0\% & 759 & 258 & 32 & 5 & 0  \\ 
    25\% & 942 & 372 & 62 & 9 & 0  \\ 
    50\% & 1594 & 799 & 155 & 22 & 0  \\ 
    $N_{N_c}^{(gauss)}$ & 2599.408 & 372.738 & 22.117 & 0.519 & 0.005  \\ 
  \end{tabular}}\\
\subfloat[$k=0.025$]{\label{tbl:nbTraj_tbl_b}
  \begin{tabular}{l|ccccc}
    Overlap & $\sigma$ & $2\sigma$ & $3\sigma$ & $4\sigma$ & $5\sigma$ \\
    \hline
    0\% & 305 & 179 & 52 & 5 & 0  \\ 
    25\% & 451 & 310 & 144 & 39 & 4  \\ 
    50\% & 1019 & 834 & 521 & 198 & 27  \\ 
    $N_{N_c}^{(gauss)}$ & 2599.408 & 372.738 & 22.117 & 0.519 & 0.005  \\ 
  \end{tabular}}\\
\subfloat[$k=0.03$]{\label{tbl:nbTraj_tbl_c}
  \begin{tabular}{l|ccccc}
    Overlap & $\sigma$ & $2\sigma$ & $3\sigma$ & $4\sigma$ & $5\sigma$ \\
    \hline
    0\% & 100 & 80 & 33 & 8 & 1  \\ 
    25\% & 186 & 166 & 110 & 46 & 7  \\ 
    50\% & 539 & 510 & 391 & 205 & 36  \\ 
    $N_{N_c}^{(gauss)}$ & 2599.408 & 372.738 & 22.117 & 0.519 & 0.005  \\ 
  \end{tabular}}
\caption{Number of independent trajectories corresponding to fluctuations $F'_{T_a} \geq a$ with a=$\sigma,2\sigma...$ in the reconstructed ensemble, for the three experiments $k=0.02,0.025,0.03$. Parameters of the \ac{gktl} experiments are $N_c = 16384$, $T_a=10\tau_c$, $\tau = \tau_c /2$ and $\epsilon = 0.02$. In each case we count as independent trajectories with an overlap of $0\%$ (zero overlap), $25\%$ and $50\%$, respectively. In addition, $N_{N_c}^{(gauss)}$ is the average number of fluctuations $F'_{T_a} \geq a$ on can expect from $N_c$ realisations obtained from a direct sampling approach. This number is based on the fact that the \ac{pdf} of $F_{T_a}$ is Gaussian. It is defined in~\eqref{eq:gaussian_number}.}
\label{tbl:nbTraj_tbl}
\end{table}
Table~\ref{tbl:nbTraj_tbl} displays the number of independent trajectories that correspond to fluctuations $F'_{T_a} \geq a$ with $a=\sigma_{T_a}, 2\sigma_{T_a},3\sigma_{T_a}...$, for each of the three experiments.
In each case, we count the number of trajectories according to the allowed degree of overlap.
For instance, allowing a $25\%$ overlap, we consider trajectories that overlap over $[0;T_a/4]$ as independent.
In order to compare with a direct sampling approach, we could perform a very long simulation of computational cost $N_c\times T_a$, divide it into $N_c$ realisations $F_{T_a}$ and count the number of fluctuations $F'_{T_a} \geq a$ above a given threshold value $a$.
However, we found that the \ac{pdf} describing the statistics of fluctuations $F'_{T_a}$---with $T_a = 10\tau_c$--- is very close to a Gaussian \ac{pdf}.
As a result, the expected number of fluctuations $F'_{T_a} \geq a$ among $N_c$ realisations is
\begin{equation}
  N^{(gauss)}_{N_c} \approx N_c \times \set{P}(F'_{T_a} \geq a) = N_c \times \frac{1}{\sqrt{2\pi \sigma_{T_a}}}\int_{a}^{\infty}e^{-x^2/2\sigma^2_{T_a}}\mathrm{d}x.
  \label{eq:gaussian_number}
\end{equation}
Considering small fluctuations $F'_{T_a} \geq a$ with $a = \sigma$ and $a = 2\sigma$, one can see in table~\ref{tbl:nbTraj_tbl} that the number of sampled independent trajectories actually decreases as $k$ is increased. This results from the fact that the overall number of independent trajectories decreases as $k$ is increased, as illustrated in figure~\ref{fig:nbIndep}.
More importantly, considering fluctuations above $3\sigma$, the \ac{gktl} experiments for $k=0.025$ and $k=0.03$ provide a larger number of fluctuations than a direct sampling approach with similar computational cost.
For instance, table~\ref{tbl:nbTraj_tbl_b} shows that the experiment with $k=0.025$ provides $5$ fully independent trajectories corresponding to fluctuations $F'_{T_a} \geq 3\sigma$.
This is ten times more than what can be expected from a direct sampling.
Notice however how drastic is the reduction of the number of independent trajectories as the allowed degree of overlap is diminished.
Allowing for a $50\%$ overlap, the reconstructed ensemble for the experiment $k=0.025$ contains roughly $200$ independent trajectories corresponding to fluctuations $F'_{T_a} \geq 3\sigma$.
Generally speaking, we observe in all three experiments that the number of independent trajectories is very small compared to the total number of trajectories $N_c = 16384$.
For instance, table~\ref{tbl:nbTraj_tbl_c} shows that, allowing a $50\%$ overlap between independent trajectories, $521$ trajectories can be used to study the dynamics leading to fluctuations $F'_{T_a} \geq 3\sigma$.
This represents $521 / N_c \approx 3\%$ of the overall computational cost of the experiment.
We stress that, in spite of this rather poor yield, the \ac{gktl} still results in a larger amount of extreme fluctuations, with respect to a direct sampling approach.
However, there is clearly room from improvement.

>>>>

Figure~\ref{fig:instant_256_20} can be better understood by visualizing the set of the last resampled trajectories. 
One can see that, ultimately, trajectories have been resampled from the same initial trajectory.
We explain this phenonemon as follows.
Because the flow dynamics hae a memory, it takes a certain time before a resampled trajectory separate from its parent.
This is clearly visible on figure~\ref{fig:instant_256_20}.
In our case, this memory originates from the deterministic nature of the dynamics, which implies that two trajectories starting from infinitesimally close starting point will overlap over a timescale called
the <<Lyapunov timesacle>>~\cite{lyapunovRef}.
However, we observe that the duration of extreme drag fluctuations is typically 5 times smaller that the Lyapunov timescale of the drag proces.
This means that the resampling of a trajectory based on the value of the drag at $t^{\star}$ cannot lead to larger drag values for $t^{\star} \leq t \leq \tau_L$.
For $t > \tau_L$, the drag process has no memory of the drag fluctuations on which the resampling was based and the probability of observing an extreme fluctuation over $\tau_l \leq t \leq T_a$ is very low.
The difference between the typical duration of drag fluctuations $\tau_c$ and the Lyapunov timescale $\tau_0$ is linked to the convective nature of the flow.
Indeed, fluid structures responsible for a particular value of the drag on the obstacle at a time $t^{\star}$ are swept away by the mean flow over a timescale $\tau_c$, shorter than the time it takes for
trajectories to separate.

@@

We performed three experiments, corresponding to three different values of the parameter $k$.
This parameter sets the amplitude of the typical fluctuations within the biased ensemble sampled by the algorithm.
In other words, a higher value of $k$ will give a higher weight to higher fluctuations, thus driving the sampling to more extreme events.
In general the value of $k$ must be set empirically, unless a \textit{large deviation regime} $T_a \to \infty$ is verified~\cite{ref_ld_regime}.
In the latter case, the parameter $k$ can be related to the typical amplitude of the sampled flucutations through the Gartner-Ellis theorem~\cite{GEref}.
In the large time limit $T_a\to \infty$, the typical level of fluctuations $\mathbb{E}_{\mathbb{P}_k}[ F_{T_a} ]$ within the biased ensemble of trajectories
can be related to the \textit{Scaled Cumulant Generating Function} $\lambda(k)$~\cite{ref}.

Figure~\ref{fig:illustration_IS} illustrates importance sampling achieved with the \ac{gktl} algorithm for the time-averaged drag $F_T$ with $T = T_a = 10\tau_c$.
For each three experiment, it displays the estimate of the biased \ac{pdf}
\begin{equation}
  \label{eq:estimate_biased_measure}
  \rho_k(F) = \mathbb{E}_{\set{P}_k}[\delta(F_{T_a}[\{\mathbf{x}\}_{0\leq t \leq T_a}]-F)]
  \approx \frac{1}{N_c}\sum_{j=1}^{N_c}\delta(F^{(j)}_{T_{a}}[\{\mathbf{x}\}_{0\leq t \leq T_a}] - F),
\end{equation}
computed over the set of $N=16384$ trajectories sampled by the algorithm.
Moreover, figure~\ref{fig:illustration_IS_drag} displays an estimate of the <<unbiased>> \ac{pdf} describing the time-averaged drag process.
It corresponds to equation~\eqref{eq:estimate_biased_measure} with $k=0$.

% Limitations

For a fixed number of trajectories $N$, there is a limited range of $k$ values for which the algorithm is able to yield an accurate estimate of $\rho_k$.
Setting $k$ to a high value causes finite population effects: trajectories overlap over most of their duration.
This effect is illustrated in figure~\ref{fig:illustration_IS}, where the histogram corresponding to the largest value of $k$ is peaked around a single value.
This is a consequence of the fact that most of the trajectories in the biased ensemble overlap, therefore leading to the same value for the averaged drag $F_T$.
See~\cite{ref1,ref2,ref3} for further details about the scaling of the required population size as $k$ is increased.

% Malgré les limitations, c'est toujours mieux
% qu'un échantillonnage direct
Unsurprisingly, sampling larger fluctuations thus reauire a larger number of populations, and therefore a larger computational cost.
Even though, table~\ref{tbl:nbTraj_tbl} shows that the \ac{gktl} algorithm is able to sample trajectories corresponding to values of the time-averaged drag that could
not be sampled through a direct sampling with equal computational cost.
Table~\ref{tbl:nbTraj_tbl} displays the number of trajectories that correspond to a time-average $F_{T_a} \geq a$, for several values of $a$.
As a comparison, it also displays the average number of such fluctuations in a long timeseries of duration $T_{tot}=10^6\tau_c$, divided in $T_{tot} / T_a$ events.
This number is computed based on the assumption that the \ac{pdf} describing the random variable $F_T$ with $T=10\tau_c$ is Gaussian, motivated by figure~\ref{fig:pdf_averared}.
While this assumption does not hold for $F_T \geq 4\sigma_T$, it may still provide numbers relevant to the comparison with the algorithm.

\begin{table}
  \centering
  \begin{tabular}{l|ccccc}
     & $\sigma$ & $2\sigma$ & $3\sigma$ & $4\sigma$ & $5\sigma$ \\
    \hline
    $k=0.02$ & 1594 & 799 & 155 & 22 & 0  \\ 
    $k=0.025$ 1019 & 834 & 521 & 198 & 27  \\ 
    $k=0.03$ & 539 & 510 & 391 & 205 & 36  \\ 
    $\mathcal{N}_{direct}$ & 2599.408 & 372.738 & 22.117 & 0.519 & 0.005  \\ 
  \end{tabular}
  \caption{Number of independent corresponding to fluctuations $F_{T_a} \geq a$ with a=$\sigma,2\sigma...$ in the biased ensemble, for the three experiments $k=0.02,0.025,0.03$. Parameters of the \ac{gktl} experiments are $N_c = 16384$, $T_a=10\tau_c$, $\tau = \tau_c /2$. Within the biased ensemble, trajectories are considered distinct if they overlap for less than $50\%$ of their duration.
    In addition, $\mathcal{N}_{direct}$ is the average number of fluctuations $F'_{T_a} \geq a$ on can expect from $N_c$ independant realisations of $F_{T_a}$, \textit{i.e.} with $k=0$.
    This number is based on the fact that the \ac{pdf} of $F_{T_a}$ is Gaussian.}
\label{tbl:nbTraj_tbl}
\end{table}
